---
title: "1.choose_data"
author: "abbey yatsko"
date: "26/08/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Choose data - choose target deadwood samples that are broadly represented across decay classes 

goals for sample analysis: 
* carbon gas flux (LGR chamber measurements)
* microbe community composition (sawdust, CTAB, then sequencing)
* moisture content
* wood chemistry or lignin percentages (?)

## Set workspace 
```{r libraries}
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Load in data and prep
```{r}
# read in full data from compiled deadwood survey
dw_survey <- read.csv("/Users/abbeyyatsko/Desktop/repos/serc_flux/data/summer2022/cleaned_survey_14-21.csv")

# from df.new: want PIECETAG, SPCODE, location information, BDS.2021 (to differentiate snag/log), and information for DC.2021 
dw_survey <- dw_survey %>%
  select(c('STEMTAG', 'PIECETAG', "QUADNAME", "QX.x", "QY.x", 'SPCODE', 'DC.2021', 'BDS.2021', 'DBH.2014', 'DBH.2017', 'DBH.2019', 'DBH.2021', 'PIECETAG', 'LENGTH.2021', 'LENGTH.2017', 'LENGTH.2014'))

# exclude samples with no information for DC.2021
# we only want to choose samples that were recorded for decay class in 2021 - most recent estimate on how decayed they are
dw_2021 <-dw_survey[!is.na(dw_survey$DC.2021),]

# filter out all snags - this study only focuses on downed deadwood 
dw_2021 <-dw_2021[!(dw_2021$BDS.2021 == 'Snag'),]
```

## Data exploration 
```{r data exploration}
# what kind of species are there?
unique(dw_2021$SPCODE)

# what is the distribution of species counts? show pecies and frequency of dw pieces in 2021 survey 
dw_2021 %>% 
  drop_na(SPCODE) %>%
  ggplot(aes(x = SPCODE))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# potential candidates:
# tulip poplar (LITU) 
# oaks (QUFA, QUERC, QUAL)
# red maple (ACRU)
# american beech (FAGR)
# sweetgum (LIST2)

# NOTE that FRPE has high hits, but this is probably because of the ash dieback that is pretty widespread 
# in the marsh region of the plot 
ash <- subset(dw_2021, SPCODE == "FRPE")

# there are mainly 1s and 2s, some 3s, a handful of 4s, and like a single 5s im pretty sure
ashplot <- ggplot(data = ash, aes(x = DC.2021)) + 
  geom_bar(position="stack", stat="count")
```

## Build sampling data
* the goal is to choose deadwood pieces that are well represented across the decay trajectory
* chosen deadwood pieces will be marked in the ForestGEO plot (18 JULY)
* these deadwood pieces will be sampled for small wood blocks, which respiration will be measured on (1 AUGUST, 8 AUGUST)

```{r build sampling data}
# look at decay class distribution by species 
ggplot(data = dw_2021, aes(x = DC.2021)) + 
  geom_bar(position="stack", stat="count")+ 
  facet_wrap(~SPCODE)+
  theme_light()

# species with the broadest distribution across 5 decay classes: 
# ACRU - red maple
# CATO6 - hickory
# COFL2 - flowering dogwood
# LIST2 - sweetgum
# LITU - tulip tree
# QUXX collectively - all of the oaks 
#   QUAL, QUCO2, QUERC, QUFA, QUPA2, QURU, QUVE

# create vector with targeted species to sample 
selected_species <- c("ACRU" , "CATO6" , "COFL2" , "LIST2" , "LITU" , "QUAL", "QUCO2", "QUERC", "QUFA", "QUPA2", "QURU", "QUVE")

# subset out data that belongs to selected_species vector
dw_samples <- dw_2021[dw_2021$SPCODE %in% selected_species, ]
dw_samples$SPCODE_all <- dw_samples$SPCODE

# look at decay class distribution by species, add counts of DW pieces:
ggplot(data = dw_samples, aes(x = DC.2021)) + 
  geom_bar(position="stack", stat="count")+ 
  facet_wrap(~SPCODE)+
  theme_light()+ 
  geom_text(aes(label = ..count..), stat = "count", position = "fill")

# create vector that contains all of the oak species, going to recategorize more generally as quercus spp. 
oak <- c("QUAL", "QUCO2", "QUERC", "QUFA", "QUPA2", "QURU", "QUVE")
dw_samples$SPCODE[dw_samples$SPCODE %in% oak] <- "QUERCUS"
dw_samples <- dw_samples %>% relocate(SPCODE_all, .before = SPCODE)

# re-generate species-level decay class distribution graph: 
ggplot(data = dw_samples, aes(x = DC.2021)) + 
  geom_bar(position="stack", stat="count")+ 
  facet_wrap(~SPCODE)+
  theme_light()+ 
  geom_text(aes(label = ..count..), stat = "count", position = "fill")

# decay class distribution graph without faceting by species: 
ggplot(data = dw_samples, aes(x = DC.2021)) + 
  geom_bar(position="stack", stat="count")+ 
  theme_light()+ 
  geom_text(aes(label = ..count..), stat = "count", position = "fill")

# these are the targeted deadwood pieces for July/August field campaign! 

# export .csv as the main sampling data sheet
# write.csv(dw_samples,"/Users/abbeyyatsko/Desktop/repos/serc_flux/data/summer2022/dw_targetsamples_presample.csv", row.names = FALSE)
```

## Notes
* some of the samples in draft_dw_samples.csv were missing coordinate info. I went in and manually added some (but not all) missing info
* additionally a field note column ('tag_notes') was added after 18 July fieldwork 
* tag_notes may require additional information to be added (lengths, diameters, identified pieces that were no longer around)
* additional information was manually entered, referencing other living/deadwood surveys in the plot