---
title: "LGR_processing_template_pilot"
author: "abbey yatsko"
date: "8/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# LGR Processing Template 
## August 2022

## PART 1 - INTRODUCTION ----
### The purpose of this document is to process and visualize gas flux data from the Los Gatos Research Ultraportable Greenhouse Gas Analyzer (LGR UGGA). Gas flux data comes from pieces of deadwood at different stages in the decay trajectory. This document uses pilot data collected from 257 samples of deadwood at the Smithsonian Environmental Research Center (SERC) in Edgewater, MD. All code is adapted from Genevieve Noyce.  

## PART 2 - PREPROCESS ----
### Set workspace, load in raw data (.txt files taken directly from the LGR include necessary packages and call on libraries

```{r workspace}
library(data.table)
library(scales)
library(ggplot2)
library(ggpubr)
library(sqldf)
library(dplyr)

# data load and prep 
# set working directory
setwd("/Users/abbeyyatsko/Downloads/")

# get names of all LGR files with GHG concentration, a.k.a the data with the form ('...f####.txt')
filenames <- list.files(pattern='f0000',full.names=T)

# read in LGR datafiles to list and go from .txt to .csv 
data <- lapply(filenames,read.csv,skip=1)

# combine all files into single dataframe for 9 days of sampling
data <- do.call(rbind,data)
```

formatting dates and times: 
```{r date/time}
# pull out date and time data
date_time <- strptime(data[,1],format='%m/%d/%Y %H:%M:%S')

# add year,month,day,JD,hour,min,sec columns to dataframe
Year <- as.numeric(format(date_time,'%Y'))
Month <- as.numeric(format(date_time,'%m'))
Day <- as.numeric(format(date_time,'%d'))
fDOY <- as.numeric(julian(date_time,'2022-01-01'))  #Change for year
Hour <- as.numeric(format(date_time,'%k'))
Min <- as.numeric(format(date_time,'%M'))
Sec <- as.numeric(format(date_time,'%S'))
data <- cbind(date_time,Year,Month,Day,fDOY,Hour,Min,Sec,data[,-1])
# save LGR data as data.table
data <- data.table(data)
```

now what we have is some better-formatted raw data that came straight from the LGR itself. A good checkpoint is to quickly plot this so that we can see that there is data for each sample day (there should be 9 total). 

preliminary viz: 
```{r prelimplots}
# for co2
co2 <- ggplot(data, aes(date_time, X.CO2.d_ppm)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("CO2 flux")

# for ch4 
ch4 <- ggplot(data, aes(date_time, X.CH4.d_ppm)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("CH4 flux")

# view co2 and ch4 plots together
ggarrange(co2, ch4)
```

## PART 3 - MERGE WITH METADATA ----
here, the goal is to take information about each individual sample that was measured for GHG flux and merge it with the flux readout that we just preprocessed. ultimately, we want to say for a given sample, here are all of the timepoints for which GHGs were measured over. then we can extract this information and have the change in GHGs through time (4 min sampling period) for each piece of deadwood that we sampled.

read in and format metadata, focusing on fraction Day of Year (fDOY):
```{r format metadata}
# day 1 data 
# read in file with metadata
meta_dat <- read.csv('/Users/abbeyyatsko/Downloads/dw_sample_cubes_all - Sheet1 (6).csv')
as.Date(meta_dat$measure_day, '%m/%d/%Y')

## Pull out date and time data
date_time_start <- strptime(paste(meta_dat$measure_day,meta_dat$resp_start),'%m/%d/%Y %H:%M:%S')

## Add 30 seconds to set buffered start time
date_time_start <- date_time_start+30

## Add 3.5 min (210 sec) to determine end time
date_time_end <- date_time_start+210

## Add fDOY columns for start and endtime to dataframe
fDOY_start <- as.numeric(julian(date_time_start,'2022-01-01'))  #Change for year
fDOY_end <- as.numeric(julian(date_time_end,'2022-01-01'))  #Change for year
meta_dat <- cbind(fDOY_start,fDOY_end,meta_dat)

# calculate average respiration temperature - add column
meta_dat$avg_respT <- rowMeans(meta_dat[ , c(14:18)], na.rm = TRUE)

# create new identifier: PIECETAG_DC (includes piecetag but also decay class info)
# this is necessary because sometimes multiple samples were taken per log (and share the same PIECETAG)
meta_dat$PIECETAG_DC <- paste(meta_dat$PIECETAG, meta_dat$decay_class, sep = "-")

# change piecetags to factor level variables 
meta_dat$PIECETAG_DC <- as.factor(meta_dat$PIECETAG_DC) 

# rename co2 and ch4 columns because naming format including '.' messes up the sql code 
data <- data %>%
  rename(CH4 = X.CH4._ppm,
         CO2 = X.CO2._ppm)

data <- data %>%
  rename(CH4_dry = X.CH4.d_ppm,
         CO2_dry = X.CO2.d_ppm)
```

merge LGR files to plot with metadata - the ultimate goal is to clip out samples and remove measurements for when the chamber was empty 
```{r sample clip}

# complete merge for respiration data and metadata
data_merge <- sqldf("select data.fDOY, data.CO2_dry, data.CH4_dry, meta_dat.PIECETAG_DC, meta_dat.avg_respT, meta_dat.species, meta_dat.moisture_percentage_g, meta_dat.decay_class, meta_dat.fDOY_start 
from data LEFT JOIN meta_dat ON (data.fDOY BETWEEN meta_dat.fDOY_start AND meta_dat.fDOY_end)")

# NOW that things are in working order and paired up: parse through only the data recorded for ch4 where PIECETAG is recorded - this means that the data actually means something and is assigned to an actual sample

# use 'complete.cases'
data_merge <- data_merge[complete.cases(data_merge), ]

# now visually check clipped bits of data 
# co2 (xlim=c(212.4705, 212.5215) corresponds to only the first day of flux measurement)
par(mfrow=c(2,1),mar=c(4,4,1,1))
with(data,plot(fDOY,CO2_dry,ylim=c(400,1300), xlim=c(212.4705, 212.5215)))
with(data_merge,plot(fDOY,CO2_dry,ylim=c(400,1300), xlim=c(212.4705, 212.5215)))

# ch4 
par(mfrow=c(2,1),mar=c(4,4,1,1))
with(data,plot(fDOY,CH4_dry,ylim=c(1.95,2.4), xlim=c(212.4705, 212.5215)))
with(data_merge,plot(fDOY,CH4_dry,ylim=c(1.95,2.4), xlim=c(212.4705, 212.5215)))
```

since the data is now merged, we can export the .csv file as an intermediate and do some last minute edits/additions to the file that will come into play in later steps

Add 'flag' columns (skip for now?) and export merged data to .csv file
```{r export data}
# d1query$Flag_CH4='Y'
# d1query$Flag_CO2='Y'
# d1query$CH4_notes=''
# d1query$CO2_notes=''
# d2query$Flag_CH4='Y'
# d2query$Flag_CO2='Y'
# d2query$CH4_notes=''
# d2query$CO2_notes=''

# export merged file (this is really clipped data, each chamber measurement corresponds to a PIECETAG_DC sample)
# write.csv(data_merge,"/Users/abbeyyatsko/Downloads/data_merged.csv", row.names = FALSE)
```

## PART 4 - PROCESS ----
### create linear plots using (fDOY ~ CO2/CH4) and then further seperating out individual records by PIECETAG. 

first, take the merged data for the fluxes and join info for samples (pilot_raw_data.csv) based on 'PIECETAG'
```{r}
# load in merged data
data_merge <- read.csv("/Users/abbeyyatsko/Downloads/data_merged.csv")

# check structure and adjust factor variables
str(data_merge)
data_merge$PIECETAG_DC <- as.factor(data_merge$PIECETAG_DC)
data_merge$species <- as.factor(data_merge$species)
data_merge$decay_class <- as.factor(data_merge$decay_class)
```

to start out by visualizing each individual sample fluxes for both co2 and ch4: 
```{r graph seperate samples}
# only graph 1st day of measurements (17 samples)
day1 <- meta_dat$PIECETAG_DC[1:17]
data_merge_day1 <- data_merge[data_merge$PIECETAG_DC %in% day1,]

co2 <- ggplot(data_merge_day1, aes(fDOY, CO2_dry, color = decay_class)) + 
  geom_point() + 
  facet_wrap( ~ PIECETAG_DC, ncol = 3, scales = "free") + 
  stat_smooth(method = "lm") +
  theme_classic()

ch4 <- ggplot(data_merge_day1, aes(fDOY, CH4_dry, color = decay_class)) + 
  geom_point() + 
  facet_wrap( ~ PIECETAG_DC, ncol = 3, scales = "free") + 
  stat_smooth(method = "lm") +
  theme_classic()
```

convert ppm to moles for CH4 and CO2
```{r}
# use known volume of chamber (in m3) to convert ppm of gas to liters of gas
# little mason jars are 4oz = 0.118294L 
# add chamber volume column: 
data_merge$chamber_vol <- 0.118294

# need to subtract volume of wood block itself: modeled as 3x3x3cm cube = 0.009 L 
# **this assumption might change: idea to model some as soil, some as the cube? to be determined**
# add sample volume column: 
data_merge$sample_vol <- 0.009

# add gas volume column 
data_merge$gas_vol <- data_merge$chamber_vol - data_merge$sample_vol

## FIX UNITS HERE - steps above are in L when they shoudl be in ml 
# CH4 
data_merge$CH4_dry_L=
  # parts CH4 per million parts air * volume of air in chamber (m3) * 1000 L per m3
  (data_merge$CH4_dry/1000000) * data_merge$gas_vol * 1000 

# CO2
data_merge$CO2_dry_L=
  # parts CO2 per million parts air * volume of air in chamber (m3) * 1000 L per m3
  (data_merge$CO2_dry/1000000) * data_merge$gas_vol * 1000 

# Use ideal gas law to calculate umol of CH4 or mmol of CO2
# CH4
data_merge$CH4_dry_umol=
  # (atm pressure * L CH4) / (R in L*atm/?K*mol * ?K temp) * 10^6 umol/mol
  ((1*data_merge$CH4_dry_L)/(0.08206*(data_merge$avg_respT+273)))*10^6

# CO2
data_merge$CO2_dry_mmol=
  # (atm pressure * L CO2) / (R in L*atm/?K*mol * ?K temp) * 10^3 mmol/mol
  ((1*data_merge$CO2_dry_L)/(0.08206*(data_merge$avg_respT+273)))*10^3

```

calculate chamber fluxes 
```{r}
## Identify start of fluxes
Time_start <- meta_dat[,c("fDOY_start","PIECETAG_DC")]
flux.times=unique(Time_start$fDOY_start)

# Create new dataframes to hold final fluxes 
fluxes.CH4=data.frame(matrix(NA,nrow=length(flux.times)))
fluxes.CO2=data.frame(matrix(NA,nrow=length(flux.times)))

## Add named columns
# CH4
fluxes.CH4$Time_start=flux.times
fluxes.CH4$flux.CH4=0
fluxes.CH4$R2.CH4=0
fluxes.CH4$p.CH4=0
# CO2
fluxes.CO2$Time_start=flux.times
fluxes.CO2$flux.CO2=0
fluxes.CO2$R2.CO2=0
fluxes.CO2$p.CO2=0

## Remove initial empty column
fluxes.CO2=fluxes.CO2[,-1]
fluxes.CH4=fluxes.CH4[,-1]

## For each start time
for (i in flux.times) {
  ## CH4 ##
  # Subset data for one chamber measurement
  temp1=subset(data_merge,fDOY_start==i)
  # Set corresponding row of output table
  j=which(flux.times==i)
  # Determine if start time has a CH4 flux
  if (nrow(temp1)>0) {
    # If so:  
    # Calulate flux in umol/day using linear regression
    mod=with(temp1,lm(CH4_dry_umol~fDOY))
    # Save flux rate and R2 and p-value of slope in corresponding row of dataframe
    # flux rate, converted from umol/day to umol/m2/day
    fluxes.CH4$flux.CH4[j]=coef(mod)[2]
    # R2 of slope
    fluxes.CH4$R2.CH4[j]=summary(mod)$r.squared
    # p-value of slope
    fluxes.CH4$p.CH4[j]=summary(mod)$coefficients[2,4]
    # If not:
    # Fill rows of table with NA    
  } else {
    fluxes.CH4$flux.CH4[j]=NA
    fluxes.CH4$R2.CH4[j]=NA
    fluxes.CH4$p.CH4[j]=NA
  }
  ## CO2 ##
  # Subset data for one chamber measurement
  temp2=subset(data_merge,fDOY_start==i)
  # Calulate flux in mol/day using linear regression
  mod=with(temp2,lm(CO2_dry_mmol~fDOY))
  # Save flux rate and R2 and p-value of slope in corresponding row of dataframe
  # flux rate, converted from mol/day to mol/m2/day
  fluxes.CO2$flux.CO2[j]=coef(mod)[2]
  # R2 of slope
  fluxes.CO2$R2.CO2[j]=summary(mod)$r.squared
  # p-value of slope
  fluxes.CO2$p.CO2[j]=summary(mod)$coefficients[2,4]
}

# merge metadata back in to flux files 
# CO2
meta_dat <- meta_dat %>%
  rename(Time_start = fDOY_start)

CO2_fluxfinal <- merge(fluxes.CO2, meta_dat, by="Time_start")

# CH4
CH4_fluxfinal <- merge(fluxes.CH4, meta_dat, by="Time_start")

## Export fluxes as .csv file
# CO2 (units are mmol/day)
write.csv(CO2_fluxfinal,"/Users/abbeyyatsko/Downloads/CO2_fluxfinal.csv", row.names = FALSE)

# CH4 (units are umol/day)
write.csv(CH4_fluxfinal,"/Users/abbeyyatsko/Downloads/CH4_fluxfinal.csv", row.names = FALSE)
```

visualize chamber fluxes 
```{r}
# decay class by CO2 flux (boxplot)
a <- ggplot(data = CO2_fluxfinal, aes(x = as.factor(decay_class), y = flux.CO2)) + 
  geom_boxplot()+ 
  geom_jitter()+
  theme_light()

# decay class by CH4 flux (boxplot)
b <- ggplot(data = CH4_fluxfinal, aes(x = as.factor(decay_class), y = flux.CH4)) + 
  geom_boxplot()+ 
  geom_jitter()+
  theme_light()

# facet by species
# CO2
a + facet_wrap(~ species, ncol=2)

# CH4
b + facet_wrap(~ species, ncol=2)

# CO2 respiration by moisture content 
c <- ggplot(data = CO2_fluxfinal, aes(x = moisture_percentage_g, y = flux.CO2)) + 
  geom_point()+ 
  theme_light()

# facet by species
c + facet_wrap(~ species, ncol=2)

# CH4 respiration by moisture content 
d <- ggplot(data = CH4_fluxfinal, aes(x = moisture_percentage_g, y = flux.CH4)) + 
  geom_point()+ 
  theme_light()

# # facet by species
d + facet_wrap(~ species, ncol=2)
```
